---
title: "Tidycensus"
output: html_document
---


# Load packages

```{r setup, include=FALSE}
rm(list=ls())

knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(echo=FALSE, results='hide', warning=FALSE, message=FALSE)

library(tidycensus)
library(tidyverse)
library(sf)
library(viridis)

if(!require("wordcountaddin")) remotes::install_github("benmarwick/wordcountaddin")

```


# Tidycensus

[Tidycensus](https://walker-data.com/tidycensus/) is an R package that allows users to interface with a select number of the US Census Bureau’s data APIs and return tidyverse-ready data frames, optionally with simple feature geometry included. 

The first step to start working with Tidycensus is to request a Census API key. 
(1) You should request an API key here: http://api.census.gov/data/key_signup.html
(2) Then, you should run the code chunk below with the census_api_key() function ONCE. Make sure to substitute the XXXXXX string for your own Census API key.
(3) You will be prompted to restart your R session. Do it, and then you can comment out the line with the census_api_key().


```{r tidycensus, include=FALSE}
#census_api_key("XXXXXXXXXXXXXXXXXXXXXXXXXXX", install = TRUE) # Substitute the XXXXXX string for your own Census API key
```


```{r tidycensus options, include=FALSE}
options(tigris_use_cache = TRUE)
```


## Searching for variables

Getting variables from the US American Community Survey (ACS) 5-Year Data (2018-2022) requires knowing the variable ID - and there are thousands of these IDs across the different files. To rapidly search for variables, use the `load_variables()` function. 
The function takes two required arguments: the year of the Census or endyear of the ACS sample, and the dataset name, which varies in availability by year. 
For this assignment, use `2022` and `acs5`.


```{r}
view_vars <- load_variables(2022, "acs5", cache = TRUE)

View(view_vars)

```

You can then filter the dataframe to focus on variables you are interested in.
The code below filters the variables with the word "income" in the concept column, which are avaiable at the Census tract level.
We add `(?i)` at the beginning of the pattern to make the search case-insensitive.

```{r}

view_vars_filter <- view_vars %>% 
  filter(str_detect(concept, "(?i)income")) %>%
  filter(geography == "tract")

View(view_vars_filter)

```

Adding more filters:

```{r}

view_vars_filter <- view_vars %>% 
  filter(str_detect(concept, "(?i)income")) %>%
  filter(str_detect(concept, "(?i)household")) %>%
  filter(str_detect(concept, "(?i)median")) %>%
  filter(geography == "tract")

View(view_vars_filter)

```


# Accessing *variables* from the ACS

The following example demonstrates how to extract a variable summarising median household income from the 2018-2022 ACS for Census tracts on the island of Manhattan (New York City), which corresponds to New York County:
```{r nyc}

nyc <- get_acs(
  state = "NY",
  county = "New York",
  geography = "tract",
  variables = "B19013_001",
  geometry = TRUE,
  year = 2022
)

head(nyc)
```

We can then map median household income in Manhattan:
```{r}
nyc %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA) + 
  scale_fill_viridis_c(option = "magma") 
```


You can replicate this process for any other variable of interest, or for any other county in the US.

In this assessment we will focus on Cook County, Illinois, which includes the city of Chicago. So, `state = "IL"` and `county = "Cook"`

# Accessing *tables*, which contain several variables, from the ACS

We can access all variables in a given category by using the `table` argument. For instance, to access all variables related to race in New York County, New York, we can use the following code:

```{r Population Race}
table_race <- get_acs(
  state = "NY",
  county = "New York",
  geography = "tract",
  table = "B02001",  # use table to GET all the variables in this given category
  output = "wide",
  year = 2022
)

table_race <- select(table_race,-c(NAME))
table_race <- rename(table_race, all_ppl_race = B02001_001E, black = B02001_003E)

head(table_race)

```

# Join

Imagine that, following the steps above, you have saved the variables we are interested in four different dataframes: table_v1, table_v2, table_v3, and table_v4. To plot them, you would need to join these dataframes with the original nyc object, as this is the only one for which we have also downloaded the geometry. 

To do this, we will utilize the fact that all dataframes have a common column, the GEOID. We can then use the `left_join()` function to merge the dataframes.

Note: Make sure you have obtained all data at the census tract level (geography = "tract"), otherwise, the GEOIDs of the different dataframes will not match. 

```{r eval = FALSE}

tables_all <- left_join(nyc, table_v1, by = "GEOID") %>%
  left_join(table_v2, by = "GEOID") %>%
  left_join(table_v3, by = "GEOID") %>%
  left_join(table_v4, by = "GEOID") 

head(tables_all)

# You can export your dataframe as a table:
st_write(tables_all, "PATH_TO_FILE/tables_all.csv", delete_dsn = TRUE)

# Or as geopackage that you can load into QGIS, for instance:
st_write(tables_all, 
        dsn = "PATH_TO_FILE/tables_all.gpkg",
        driver = "GPKG", 
        delete_dsn = TRUE)

```


# Word count

```{r word_count}

word_count()

```

